---
output: github_document
bibliography: references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ricci

<!-- badges: start -->

[![R-CMD-check](https://github.com/lschneiderbauer/ricci/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/lschneiderbauer/ricci/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/lschneiderbauer/ricci/graph/badge.svg)](https://app.codecov.io/gh/lschneiderbauer/ricci) [![CRAN status](https://www.r-pkg.org/badges/version/ricci)](https://CRAN.R-project.org/package=ricci) [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

<!-- badges: end -->

The goal of {ricci} is to provide a *compact*[^1] R interface for performing [tensor calculations](https://en.wikipedia.org/wiki/Ricci_calculus). This is achieved by allowing (upper and lower) index labeling of R’s `array` and making use of Ricci calculus conventions to *implicitly* trigger **contractions** and diagonal subsetting. Explicit tensor operations, such as **addition, subtraction and multiplication of tensors** via the standard operators (`*`, `+`, `-`, `/`, `==`), **raising and lowering indices**, taking **symmetric** or **antisymmetric tensor parts**, as well as the **Kronecker product** are available. Common tensors like the Kronecker delta, Levi Civita epsilon, certain metric tensors, the Christoffel symbols, the Riemann as well as Ricci tensors are provided. The **covariant derivative** of tensor fields w.r.t to any metric tensor can be evaluated. An effort was made to provide the user with meaningful error messages.

[^1]: By compact interface, we mean an interface that is concise and non-verbose. The author is of the opinion that the less we need to write in order to express an intent, the less likely are we to make mistakes.

{ricci} uses the [calculus](https://calculus.eguidotti.com/) package [@guidotti2022] behind the scenes to perform calculations and provides an alternative interface to a subset of its functionality. Notably, {[calculus](https://calculus.eguidotti.com/)} also supports symbolic calculations which also enables {ricci} to do the same. Symbolic expressions are optionally (and transparently) simplified when the [Ryacas](https://r-cas.github.io/ryacas/) package is installed.

## Installation

You can install the development version of ricci from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("lschneiderbauer/ricci")
```

## Example

The central object is R's `array`. Adding index slot labels allows us to perform common tensor operations implicitly. After the desired calculations have been carried out we can remove the labels to obtain an ordinary `array`.

The following (admittedly very artificial) example shows how to express the contraction of two tensors, and subsequent symmetrization and diagonal subsetting. For demonstration purposes we use an arbitrary array of rank 3.

```{r example_num}
library(ricci)

# numeric data
a <- array(1:(2^3), dim = c(2, 2, 2))

# create labeled array (tensor)
(a %_% .(i, j, k) * 
  # mutliply with a labeled array (tensor) and raise index i and k
  a %_% .(i, l, k) |> r(i, k, g = g_mink_cart(2))) |> 
  # * -i and +i as well as -k and +k dimension are implictely contracted
  # the result is a tensor of rank 2
  sym(j, l) |> # symmetrize over i and l
  subst(l -> j) |> # rename index and trigger diagonal subsetting
  as_a(j) # we unlabel the tensor with index order (j)
```

The same instructions work for a symbolic array:

```{r example_symb}
# symbolic data
a <- array(paste0("a", 1:(2^3)), dim = c(2, 2, 2))

(a %_% .(i, j, k) * 
  # mutliply with a labeled array (tensor) and raise index i and k
  a %_% .(i, l, k) |> r(i, k, g = g_mink_cart(2))) |> 
  # * -i and +i as well as -k and +k dimension are implictely contracted
  # the result is a tensor of rank 2
  sym(j, l) |> # symmetrize over i and l
  subst(l -> j) |> # rename index and trigger diagonal subsetting
  as_a(j) # we unlabel the tensor with index order (j)
```

For more details, see `vignette("ricci", package = "ricci")`. For more information about how to use tensor fields and the covariant derivative, see `vignette("tensor_fields", package = "ricci")`.
