---
output: github_document
bibliography: references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ricci

<!-- badges: start -->

[![R-CMD-check](https://github.com/lschneiderbauer/ricci/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/lschneiderbauer/ricci/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/lschneiderbauer/ricci/graph/badge.svg)](https://app.codecov.io/gh/lschneiderbauer/ricci) [![CRAN status](https://www.r-pkg.org/badges/version/ricci)](https://CRAN.R-project.org/package=ricci) [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

<!-- badges: end -->

The goal of {ricci} is to provide a *compact*[^1] R interface for performing [tensor calculations](https://en.wikipedia.org/wiki/Ricci_calculus). This is achieved by labeling (upper and lower) index slots of R's `array` and making use of Ricci calculus conventions to *implicitly* trigger contractions and diagonal subsetting. Explicit tensor operations, such as addition, multiplication of tensors, raising and lowering indices, or the Kronecker product are also available via the standard operators (`*`, `+`, `-`). Common tensors like the Kronecker delta, Levi Civita epsilon, and certain metric tensors are provided.

[^1]: By compact interface, we mean an interface that is concise and non-verbose. The author is of the opinion that the less we need to write in order to express an intent, the less likely are we to make mistakes.

{ricci} uses the [calculus](https://calculus.eguidotti.com/) package [@guidotti2022] behind the scenes to perform calculations and simply provides an alternative interface to a subset of its functionality.

## Installation

You can install the development version of ricci from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("lschneiderbauer/ricci")
```

## Examples

The central object is R's `array`. Adding index slot labels allows us to perform common tensor operations implicitly. After the desired calculations have been carried out we remove the labels to obtain an `array` again.

The following example shows how to express the contraction of two tensors, where one index has to be raised, and subsequent diagonal subsetting. For demonstration purposes we use an arbitrary array of rank 3.

```{r example}
library(ricci)

# a choice of metric tensor is required
# for raising / lowering indices
options(ricci.g = g_mink(2)())

# the data
a <- array(1:(2^2*3), dim = c(2,2,3))

(a %_% .(i, j, A) * # create labeled tensor
    a %_% .(j, i, A) |> .t(j -> +j)) |> # create a labeled tensor and raise index j
  # * -j and +j dimension are implictely contracted
  # * the i-diagonal is selected
  # the result is a tensor of rank 2
  .a(i, A) # we unlabel the tensor with index order (i, A)
```

Below we outline more details on possible individual tasks.

### Creating a labeled tensor

We can use the array `a` to create a labeled tensor with lower index labels i, j, and k:

$$
a_{ijk}
$$

```{r}
a %_% .(i, j, k)
```

By default, indices are assumed to be lower indices. We can use a "+" prefix to create an upper index label.

$$
a_{ij}^{\;\;k}
$$

```{r}
a %_% .(i, j, +k)
```

### Performing calculations

Creating labels on its own is not very interesting nor helpful. The act of labeling tensor index slots becomes useful when the labels are set such that they trigger implicit calculations, or they are combined with other tensors via multiplication or addition.

#### Contraction

Repeated index labels with opposite position are implicitly contracted.

$$
b_j=a_{i\;k}^{\;i}
$$

```{r}
b <- a %_% .(i, +i, k)
b

# retrieve array
b |> .a(k)
```

#### Diagonal subsetting

Repeated labels on the same position (upper or lower) will trigger diagonal subsetting.

$$
c_{ik}=a_{iik}
$$

```{r}
c <- a %_% .(i, i, k)
c

# retrieve array
c |> .a(i, k)
```

#### Outer tensor product

The same conventions apply for arbitrary tensor multiplication.

$$
d_{ijklmn}=a_{ijk}a_{lmn}
$$

```{r}
d <- a %_% .(i, j, k) * a %_% .(l, m, n)
```

#### Tensor multiplication w/ contractions

$$
e=a_{ijk}a^{ijk}
$$

```{r}
e <- a %_% .(i, j, k) * a %_% .(+i, +j, +k)
e

# convert to number
as.numeric(e)
```

#### Tensor multiplication w/ contractions and subsetting

$$
f_j=a_{ijk}a^{i\;k}_{\;j}
$$

```{r}
f <- a %_% .(i, j, k) * a %_% .(+i, j, +k)
f

# retrieve array
f |> .a(j)
```

#### Tensor addition

Tensor addition is taking care of correct index slot matching (by index labels), so the position of the index does not matter.

$$
g_{ijk} = a_{ijk} + a_{jik}
$$

```{r}
g <- a %_% .(i, j, k) + a %_% .(j, i, k)
g

g |> .a(i, j, k)
```
